---
title: "Using the afdx package to fit a logit exponential model to estimate the attributable fraction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Attributable fraction using a logitexponetial model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
By John Aponte and Orvalho Augusto.

## Introduction

In malaria endemic areas, asymptomatic carriage of malaria parasites occurs
frequently and the detection of malaria parasites in blood films from a febrile 
individual does not necessarily indicate clinical malaria. 

A case definition for symptomatic malaria that is used widely in endemic areas 
requires the presence of fever or history of fever together with a parasite density above a 
specific cutoff. If the parasite density is equal or higher than the cutoff point, 
the fever is considered due to malaria.

How to estimate what is the sensitivity and specificity of the cutoff point in
the classification of the fever, without knowing what is the true value of the fevers due to 
malaria?

Using the attributable fraction, one can estimate the expected
number of true cases due to malaria and with the positive predictive value associated
with a given cutoff point, we can estimate the expected number of true cases among
the fever cases that have a parasite density higher or equal than the selected cutoff
point. Based on this values, the rest of the 2x2 table can be completed and the
sensitivity, specificity and negative predictive value.

In order to estimate the attributable fraction and the positive predictive value,
we follow the method proposed by Smith et al. (1) fitting a logistic exponential model.  
Here it is presented how to do it with the `afdx` package for the R-software.


## Example using synthetic data

The data used in this example (malaria_df1) is a simulated data set as seen frequently in malaria cross-sectionals 
where two main outcomes are measured, the presence of fever or history of fever
(fever column) and the measured parasite density in parasites per $\mu l$ (density column).

```{r setup}
library(afdx)
```

```{r echo = F, result = 'markup'}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(kableExtra))

kable(
  head(malaria_df1, n = 6),
       caption = "head(malaria_df1) first 6 observations", 
       format = "html") %>%
kable_styling( position = "left")
```

In this simulation, there are 2000 observations, from which `r sum(malaria_df1$fever)`
have fever or history of fever and `r sum(malaria_df1$density > 0)` have a 
density of malaria greater than 0. A total of  `r sum(malaria_df1$fever * (malaria_df1$density > 0))`
have both fever and a malaria density higher than 0.

```{r echo = F, result = 'asis'}

cutoffs <- c(0,1,100,200,400,800,1600,3200,6400,12800,25600,51200, 102400, 204800)
data <- 
  malaria_df1 %>%
  mutate(k = cut(density,c(cutoffs,Inf), include.lowest =T, labels = cutoffs)) %>%
  group_by(k,fever) %>%
  tally() %>%
  mutate(category = ifelse(fever ==1,"n (fever)","m (no fever)")) %>%
  select(-fever) %>%
  pivot_wider(names_from = "category", values_from = "n", values_fill = list(n = 0)) %>%
  rename(`k (category lower limit)` = k)

kable(data, "html", caption="Distribution of fevers by density categories") %>%
kable_styling(position = "left")  
```


## The logit exponential model

Smith et al (1) investigate different models to describe the risk of fever due 
to malaria. Given that the association is not linear, the best model found was a
logit exponential model:

$logit(\pi_i) = \beta(x_i)^\tau$

where $\pi_i$ is the probability of fever for the observation $i$ and $x_i$ is 
the parasite density for observation $i$. 

The attributable fraction $\lambda$ is estimated as:

$\lambda = (1/N)\sum{_i}{(R_i - 1)/R_i}$

where $N$ is the number of fever cases, and $R_i=exp[\beta(x_i)^\tau]$

If a case of malaria is defined as a case of fever with a malaria density equal
greater than a cutoff $c$, $n_c$ is the number of fever cases that acomplish that
definition and the proportion of these diagnosis cases that are attributable to
malaria $\lambda_c$ is estimated by

$\lambda_c = (1/n_c)\sum{_i}d_{c,i}(R_i-1)/R_i$

where $d_{c,i}$ is and indicator with value 1 if that fever case statisfies the
malaria case definition and 0 otherwise

Sensitivity for the cuttof $c$ is defined then as: $n_c\lambda_c / N\lambda$,

Specificity is defined as: $1-n_c(1-\lambda_c)/N(1-\lambda)$

Predictive positive value as: $n_c\lambda_c/n_c = \lambda_c$

Negative predicted value as: $(N(1-\lambda) - n_c(1-\lambda_c))/(N-n_c)$

```{r echo =FALSE}
fit <- fit <- logitexp(malaria_df1$fever, malaria_df1$density)
cutoff = 500
xx <- senspec(fit,cutoff)
N = sum(malaria_df1$fever)
nc = sum(malaria_df1$density > cutoff & malaria_df1$fever)
L = fit$af
lc = xx[1,"ppv"]
df <- data.frame(
  "True Malaria" = c(nc*lc, N*L-nc*lc, N*L),
  "Other aetiology" = c(nc*(1-lc), N*(1-L)- nc*(1-lc), N-nc),
  "All" = c(nc, N-nc, N)
)
row.names(df)<- c("Malaria case (fever and density > 500)", "No case", "Total")
df <- round(df,1)

```

For example, if 500 is selected as cutoff point, there are `r nc` cases of fever 
with density greater or equal than 500 from a total of `r N` fevers. From the model,
the estimated attributable fraction is `r round(L*100,2)`% and the proportion
attributed to malaria for fevers equal or greater than 500 is `r round(lc*100,2)`%

The total number of true malaria cases is estimated as $N\lambda=$ `r round(N*L,1)` and
the number of true malaria cases in those that accomplish the definition as $n_c\lambda_c$ = `r round(nc*lc,1)`.
The full 2x2 table can be filled and the corresponding sensitivity, specificity and preditive values calculated.


```{r results='markdown', echo = F}
kable(df, "html", caption = "Case definition with 500 as cutoff") %>% kable_styling(position = "left")
```

## Estimating the sensitivity and specificity

The `adfx` provide functions that facilitate the fitting of the logit exponential
model and to estimate the sensitivity, specificity, positive and negative predictive
values for different cutoff points.

```{r}

fit <- logitexp(malaria_df1$fever, malaria_df1$density)
fit

senspec(fit, c(1,100,500,1000,2000,4000,8000,16000, 32000,54000,100000))
```

## Graph of diagnostic characteristics for all cutoff points

The function `make_cutoffs()` find the densities where there is change in 
the number of positives and can be used to estimate the characteristics of all
cutoff points

```{r, fig.width = 8, fig.height=6, out.width= "100%", dpi = 300}
cutoffs <-  make_cutoffs(malaria_df1$fever, malaria_df1$density)
dxp <- senspec(fit, cutoffs[-1])
dxp %>%
  data.frame(.) %>%
  pivot_longer(-cutoff, names_to = "Indicator",values_to = "Value") %>%
  ggplot(aes(x = cutoff, y = Value, color = Indicator, linetype = Indicator)) +
  geom_line() +
  scale_x_log10("Cutoff")
```

## ROC curve

The receiver operative curve can be estimated from the sensitivity and specificity
values


```{r, fig.width = 8, fig.height=6, out.width= "100%", dpi = 300}
rocdf <-dxp %>%
  data.frame(.) %>%
  ## add the corners
  bind_rows(
    data.frame(
      sensitivity= c(1,0),
      specificity= c(0,1)
    )
  ) %>%
  # generate the 1-specificity
  mutate(`1-specificity` = 1 - specificity) 

  # make the graph
  ggplot(rocdf, aes(x = `1-specificity`, y = sensitivity)) +
  geom_line()+
  ggtitle("ROC curve")
  
  # Estimate the area under the curve
  library(DescTools)
  AUC(rocdf$`1-specificity`, rocdf$sensitivity)
  
```



## Bibliography
1. Smith T, Schellenberg JA, Hayes R. Attributable fraction estimates and case definitions for malaria in endemic areas. Stat Med. 1994 Nov 30;13(22):2345–58. 
