---
title: "Attributable fraction using logitexponetial model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Attributable fraction using logitexponetial model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

In malaria endemic areas, asymptomatic carriage of malaria parasites occurs
frequently and the detection of malaria parasites in blood films from a febrile 
individual does not necessarily indicate clinical malaria. 

A case definition for symptomatic malaria that is used widely in endemic areas 
requires the presence of fever or history of fever together with a parasite density above a 
specific cutoff. If the parasite density is equal or higher than the cut-off point, 
the fever is considered due to malaria.

How to estimate what is the sensitivity and specificity of the cut-off point in
the classification of the fever, without knowing what is the true value of the fevers due to 
malaria?

Using the attributable fraction, one can estimate the expected
number of true cases due to malaria and with the positive predictive value associated
with a given cut-off point, we can estimate the expected number of true cases among
the fever cases that have a parasite density higher or equal than the selected cut-off
point. Based on this values, the rest of the 2x2 table can be completed and the
sensitivity, specificity and negative predictive value.

In order to estimate the attributable fraction and the positive predictive value,
we follow the method proposed by Smith et al. (1) fitting a logistic exponential model.


## Example using synthetic data
```{r setup}
library(afdx)
```


```{r echo = F, result = 'markup'}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(kableExtra))

kable(
  head(malaria_df1),
       caption = "head(malaria_df1)", 
       format = "html") %>%
kable_styling( position = "left")
```

The data used in this example (malaria_df1) is a simulated data from a malaria cross-sectional 
where two main outcomes are measured, the presence of fever or history of fever
(fever column) and the measured parasite density in parasites per ul (density column).


In this example, there is a total of 2000 observations, of which `r sum(malaria_df1$fever)`
have fever or history of fever and `r sum(malaria_df1$density > 0)` have a 
density of malaria greater than 0. A total of  `r sum(malaria_df1$fever * (malaria_df1$density > 0))`
have both fever and a malaria density higher than 0.

```{r echo = F, result = 'asis'}

cutoffs <- c(0,1,100,200,400,800,1600,3200,6400,12800,25600,51200, 102400, 204800)
data <- 
  malaria_df1 %>%
  mutate(k = cut(density,c(cutoffs,Inf), include.lowest =T, labels = cutoffs)) %>%
  group_by(k,fever) %>%
  tally() %>%
  mutate(category = ifelse(fever ==1,"n (fever)","m (no fever)")) %>%
  select(-fever) %>%
  pivot_wider(names_from = "category", values_from = "n", values_fill = list(n = 0)) %>%
  rename(`k (category lower limit)` = k)

kable(data, "html", caption="Distribution of fevers and denstities") %>%
kable_styling(position = "left")  
```



## The logit exponential model

## Estimating the sensitivity and specificity

## Graph of diagnostic characteristics for all cut-off points

## ROC curve

## Bibliography